---
title: "capstoneProject_v0.2"
author: "IcedMcstuffin"
date: "2026-05-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 0. Introduction
This project will work on making a next text predictor with the utilisation of Natural Language Processing. This project uses the files named LOCALE.blogs.txt where LOCALE is the each of the four locales en_US, de_DE, ru_RU and fi_FI. The data is from a corpus called HC Corpora.

```{r message=FALSE}
library(tidytext)
library(dplyr)
```

# 1. Task1: Getting and cleaning the data
We will load the blogs, twitter and news data into a combined dataframe and prepare a sample using the rbinom function.

## 1.1 Reading the data and preparing a combined dataset
```{r warning=FALSE, eval=FALSE}
con <- file("final/en_US/en_US.blogs.txt", "r")
rawblogs <- data.frame(text = readLines(con), source = "blog")
close(con)

con <- file("final/en_US/en_US.news.txt", "r")
rawnews <- data.frame(text = readLines(con), source = "news")
close(con)

con <- file("final/en_US/en_US.twitter.txt", "r")
rawtwitter <- data.frame(text = readLines(con), source = "twitter")
close(con)

# Combine all three into one dataset with a unique id and source column
rawdata <- bind_rows(rawblogs, rawnews, rawtwitter) %>%
  mutate(id = row_number()) %>%
  select(id, source, text)
```

## 1.2 Tokenising each word in the sample dataset accounting for punctuation and numbers
```{r eval=FALSE}
tokenblogs <- rawdata %>%
  unnest_tokens(word, input = text) %>%
  filter(grepl("^[a-z]+$", word))
```

# 2. Task 2: Exploratory analysis

### 2.1 Distribution of word frequencies
```{r  eval = FALSE}
wordfrequency <- tokenblogs %>% count(word, sort = TRUE)
```

```{r echo = FALSE, cache = TRUE}
wordfrequency <- readRDS("data/wordfrequency.rds")
```


```{r fig.height=10, fig.width=20, cache = TRUE}
barplot(wordfrequency$n[1:10], names.arg = wordfrequency$word[1:10], las = 2,
        main = "Frequency of most common words", ylab = "Frequency", col = "steelblue")
```

### 2.2 Frequencies of 2-grams and 3-grams

Generating a 2-gram dataset
```{r eval = FALSE}
# 2-gram generation
bigram <- tokenblogs %>%
  group_by(id) %>%
  mutate(nextword = lead(word)) %>%
  filter(!is.na(nextword)) %>%
  mutate(bigram = paste(word, nextword))
```

Sorting the table by frequency of each 2-gram
```{r eval = FALSE}
bigramfrequency <- bigram %>%
  count(word, nextword, bigram, sort = TRUE)
```

```{r echo = FALSE, cache = TRUE}
bigramfrequency <- readRDS("data/bigramfrequency.rds")
```


```{r cache=TRUE, fig.height=10, fig.width=20}
barplot(bigramfrequency$n[1:10], names.arg = bigramfrequency$bigram[1:10], las = 2,
        main = "Frequency of most common bigrams", ylab = "Frequency", col = "steelblue")
```

Generating a 3-gram dataset
```{r eval = FALSE}
# 3-gram generation
trigram <- tokenblogs %>%
  group_by(id) %>%
  mutate(nextword = lead(word), nexttonextword = lead(word, n = 2L)) %>%
  filter(!is.na(nextword), !is.na(nexttonextword)) %>%
  mutate(trigram = paste(word, nextword, nexttonextword))
```

Sorting the table by frequency of each 3-gram
```{r eval=FALSE}
trigramfrequency <- trigram %>%
  count(word, nextword, nexttonextword, trigram, sort = TRUE)
```

```{r echo=FALSE, cache=TRUE}
trigramfrequency <- readRDS("data/trigramfrequency.rds")
```


```{r cache=TRUE, fig.height=10, fig.width=20}
barplot(trigramfrequency$n[1:10], names.arg = trigramfrequency$trigram[1:10], las = 2,
        main = "Frequency of most common trigrams", ylab = "Frequency", col = "steelblue")
```

### 2.3 How many unique words do you need in a frequency sorted dictionary to cover 50% of all word instances in the language? 90%?
```{r cache=TRUE}
cumulative <- cumsum(wordfrequency$n) / sum(wordfrequency$n)
words_50 <- which(cumulative >= 0.50)[1]
words_90 <- which(cumulative >= 0.90)[1]
cat("Unique words needed to cover 50% of all instances:", words_50, "\n")
cat("Unique words needed to cover 90% of all instances:", words_90, "\n")
```

### 2.4 How do you evaluate how many of the words come from foreign languages or profane?
Foreign language detection can be achieved by including a dictionary dataset and ensuring that every word token is included in the dictionary. The same can be established by a profanity dictionary and ensuring that every word token is not included in the dictionary.

# 3. Task 3: Modelling an ngram function based on the dataset
A Backoff model will be used where primarily a 4-gram will be assessed, and if unseen, 3-gram will be assessed, and if it falls in an unseen n-gram, we will fall to 2-gram, which if unseen, we will fall back to 1-gram. In case of 1-gram, we will return the most common word.

```{r cache = TRUE}
predictword <- function(word1, word2) {
  word1 <- tolower(word1)
  word2 <- tolower(word2)


  # Trigram check
  result <- trigramfrequency %>%
    filter(word == word1, nextword == word2) %>%
    arrange(desc(n))

  if (nrow(result) > 0) {
    return(result$nexttonextword[1:min(3, nrow(result))])
  }

  # Backoff to bigram check
  result <- bigramfrequency %>%
    filter(word == word2) %>%
    arrange(desc(n))

  if (nrow(result) > 0) {
    return(result$nextword[1:min(3, nrow(result))])
  }

  # Backoff to unigram
  return(wordfrequency$word[1:3])
}
```

### Task 4: Efficiency analysis
The time taken for this predictive model can be assessed.
```{r cache=TRUE}
system.time(predictword("will", "see"))
```
